---
title: "West Nile Notebook"
author: "Louie Dinh"
date: "June 2, 2015"
output: html_document
---

# Problem Statement #

We are trying to predict the occurrence of West Nile virus at any trap location in even years. That is the WNV variable.

# Exploratory Data Analysis #

```{r}
# Load tools
library(dplyr)
library(ggplot2)
```

```{r}
# Load the data
train_data <- read.csv("data/train.csv")
test_data <- read.csv("data/test.csv")
str(train_data)

# Clean up
train_data$Date <- as.Date(train_data$Date)
```

Out of 116293 observations, we have 551 cases of West Nile detected.
Rows are split up so that each one is unique on trap/species only if a mosquito was present.

*Observation*: WnvPresent much more likely when more mosquitos are caught.

It seems like traps with WnvPresent are different
WnvPresent:
  Average NumMosquitos:, 26.35
!WnvPresent:
  Average NumMosquitos: 12.10
```{r}
ggplot(train_data, aes(WnvPresent, NumMosquitos)) + geom_boxplot()
train_data %>% group_by(WnvPresent) %>% summarize(avg = mean(NumMosquitos))
ggplot(train_data, aes(NumMosquitos)) + facet_grid(~ WnvPresent) + geom_bar()
```

Is there a separation based on the species of mosquitos?
*Observation*: Some species of mosquitos are *much* more prevalent
*Observation*: Detection of WN has only occurred in 3 species: pipiens, restuans and restuan/pipiens hybrids

```{r}
train_data %>% group_by(Species) %>% summarize(entries=n(), wn=sum(WnvPresent == "1"), m_count=sum(NumMosquitos), p_wn=wn/entries)
```


What is the prevalence of WN based on month of the year?
```{r}
# Add month as an indicator
train_data$Month <- factor(months(train_data$Date))

ggplot(train_data, aes(WnvPresent)) + facet_grid(~Month) + geom_bar()
```


```{r, echo=FALSE}
plot(cars)
```

